Obsidian:\\A Domain Specific Embedded Language\\for Parallel Programming of Graphics Processors

### 2013-10-02 src2/arbb/paperThesis.tex
apa

\vspace{2mm}
\begin{Verbatim}[commandchars=\\\{\}]
\kw{int} main() 
{ 
  double a[SIZE];
  double b[SIZE];

  for ( \kw{int} i = 0; i < SIZE; ++i) { 
    a[i] = ...; b[i] = ...;
  }
  dense<f32> va, vb;
  f32 vc;
  bind(va, a, SIZE); 
  bind(vb, b, SIZE); 
  call(dot\_product)(va, vb, &vc); 
  ...
}
\end{Verbatim}

>>>>

\vspace{2mm}
\begin{Verbatim}[commandchars=\\\{\}]
\kw{int} main() 
\{ 
  double a[SIZE];
  double b[SIZE];

  for ( \kw{int} i = 0; i < SIZE; ++i) \{ 
    a[i] = ...; b[i] = ...;
  \}
  dense<f32> va, vb;
  f32 vc;
  bind(va, a, SIZE); 
  bind(vb, b, SIZE); 
  call(dot\_product)(va, vb, &vc); 
  ...
\}
\end{Verbatim}

### 2013-10-02 src2/hl/paperThesis.tex
apa

\documentclass[preprint]{sigplanconf}

@@

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{placeins}
\usepackage{tikz}
\usepackage{pgfplots}
\usetikzlibrary{shapes,snakes,positioning}
\usepackage{fancyvrb}
\usepackage[T1]{fontenc}
%\newcounter{arraycard}

\begin{document}

\conferenceinfo{WXYZ '05}{date, City.} 
\copyrightyear{2005} 
\copyrightdata{[to be supplied]}

\titlebanner{DRAFT}        % These are ignored unless
\preprintfooter{Obsidian: GPU kernel generation}   % 'preprint' option specified.

%\title{Embedded Language for High Performance GPU Computing}
%\title{Towards An Embedded Language\\ for High Performance GPU Computing}  
%\title{An Embedded Language for Resource Aware \\ GPU Kernel Programming}  
\title{A High-Level Embedded Language for Low-Level GPU Kernel Programming}

@@

\authorinfo{Bo Joel Svensson \and Mary Sheeran} 
           {Chalmers University of Technology}
           {\{joels,ms\} at chalmers dot se}

\maketitle
% ---------------------------------------------------------------------------
%
% --------------------------------------------------------------------------- 
\begin{abstract}

@@

\end{abstract}

%\category{CR-number}{subcategory}{third-level}
\category{D.3.2}{Programming Languages}{Language Classifications}[Applicative (functional) languages; Concurrent, distributed, and parallel languages]
\category{D.3.4}{Programming Languages}{Processors}[Code generation]

%\terms
%term1, term2
\terms
Languages, Performance

%\keywords
%keyword1, keyword2
\keywords
Data parallelism, array programming, embedded language

% ---------------------------------------------------------------------------
% INTRODUCTION 
% --------------------------------------------------------------------------- 
\section{Introduction}

@@

% ---------------------------------------------------------------------------
% Motivation and Contributions
% --------------------------------------------------------------------------- 
\subsection{Motivation and contributions}

@@

% ---------------------------------------------------------------------------
% The GPU 
% --------------------------------------------------------------------------- 
\subsection{The GPU and CUDA}
\label{sec:GPU}

@@

% ---------------------------------------------------------------------------
% PROGRAMMING IN OBSIDIAN 
% --------------------------------------------------------------------------- 
\section{A taste of Obsidian programming}

@@

There are two different array representations, {\em Pull} and 
{\em Push} arrays. A pull array naturally represents gather operations,
while a push array captures scattering operations. Reference \cite{Obsidian} 
describes the introduction of push arrays into Obsidian.

@@

% ---------------------------------------------------------------------------
% OBSIDIAN Internals 
% --------------------------------------------------------------------------- 
\section{Obsidian internals}

@@

% ---------------------------------------------------------------------------
% Elements
% --------------------------------------------------------------------------- 
\subsection{Scalars}

@@

% ---------------------------------------------------------------------------
% PUSH AND PULL ARRAYS 
% --------------------------------------------------------------------------- 
\subsection{Pull arrays}

@@

Very similar representations are used 
in many other embedded languages. The 
embedded language Pan ~\cite{ELLIJFP} used a similar representation for 
images and it has been used since in Obsidian, Feldspar ~\cite{FELDSPAR} 
and Repa ~\cite{REPA}. In Obsidian, there is a {\tt force} function that,
when applied to a pull array, writes the elements of that array 
into shared memory using as many threads as the pull array is long. Hence, 
the {\tt force} function can be used to express parallelism in Obsidian.

@@

% ---------------------------------------------------------------------------
% Thread, Block and Grid Computations 
% --------------------------------------------------------------------------- 
\subsection{Thread, Block and Grid Computations} 
\label{sec:Program}

The programs that can be specified in Obsidian
are represented in a {\tt Program} data type.
This data type also represents the lowest level programming interface 
exposed by Obsidian. In reference ~\cite{CSORT} we make use of these low-level 
capabilities to implement histograms using atomic operations (not shown here).

@@

% ---------------------------------------------------------------------------
% Push Arrays
% --------------------------------------------------------------------------- 
\subsection{Push arrays} 
\label{sec:PushArrays}

@@

% ---------------------------------------------------------------------------
% Interplay 
% --------------------------------------------------------------------------- 
\subsection{Push and pull array interplay} 
\label{sec:interplay}

@@

% ---------------------------------------------------------------------------
% Compilation to CUDA 
% --------------------------------------------------------------------------- 
\subsection{Compilation to CUDA}
\label{sec:Compile}

@@

% ---------------------------------------------------------------------------
% More example programs 
% --------------------------------------------------------------------------- 
\section{Case studies} 
\label{sec:CASESTUDIES}

@@

%% ------------------------------------------------------------------------
%% Reduction
%% ------------------------------------------------------------------------
\subsection{Reduction}  
\label{sec:redOpt}

\input{reductionfigs.tex}

@@

Some of the reduction 
kernels will also require that the operation is commutative. 
%A typical reduction is to sum up the elements of an array, here this is used 
%as the running example. In this section we implement a series of reduction 
%kernels. 
To illustrate the kind of low level control that an 
Obsidian programmer has over expressing details of a kernel, we show a series 
of reduction kernels, each with different optimisations applied. Many of the 
optimisations that are applied to the kernels are influenced by a 
tutorial presentation from NVIDIA~\cite{RedOpt}.

@@

\subsubsection{Reduction 1}

@@

\subsubsection{Reduction 2}

@@

\subsubsection{Reduction 3}

@@

\subsubsection{Reduction 4}

@@

\subsubsection{Reduction 5}

@@

\subsubsection{Reductions 6 and 7}

@@

\subsubsection{Larger local reductions} 
Figure \ref{fig:reducetablel} shows running time results for kernels that are 
very similar to the ones described above, but with a larger number 
of elements per local reduction (that is per block). This means that all the kernels need 
to introduce a small amount of sequentiality at the beginning. This is 
done by introducing a new force function:

@@

\subsection{Scan}

@@

\input{scanfigs.tex}

@@

\subsubsection{two elements per thread}

@@

Both to avoid conditionals and to allow
for larger scans per block, we move to two elements per thread.
Each phase of the algorithm is a parallel for loop that is executed 
by half as many threads as there are elements to scan. The body of the loop
performs one operation and one copy, using bit-twiddling to compute indices.
Note the use of two {\em write functions} in sequence.
Similar patterns were used in our implementations of sorting networks~\cite{Obsidian}, for similar reasons.

@@

\section{Combining kernels to solve large problems} 
\label{sec:Benchmarks}

@@

\subsection{Large reductions}

@@

% ---------------------------------------------------------------------------
% RELATED WORK 
% --------------------------------------------------------------------------- 
\subsection{Related work}

@@

While remaining in the imperative world, but going all the way to the 
other end of the high-level - low-level spectrum, we have the NVIDIA
Thrust Library~\cite{Thrust}.
Thrust offers a programming model where details of GPU architecture are 
completely abstracted away. Here, the programmer expresses algorithms 
using building blocks like: \emph{Sort}, \emph{Scan} and \emph{Reduce}.

Data.Array.Accelerate is a language embedded in Haskell for GPU programming 
~\cite{accelerate}. The abstraction level is comparable to that of Thrust. In 
other words, Accelerate hides most GPU details from the programmer. Accelerate 
provides a set of operations (that are parallel and suitable for GPU execution, 
much like in Thrust) implemented as skeletons. Recent work has permitted
the optimisation of Accelerate programs using fusion techniques to decrease
the number of kernel invocations needed (see reference ~\cite{ACCOPT}).
%Accelerate programs are being optimised using using fusion 
%techniques to decrease the number of kernel invocations needed. 
It seems to us that when using Accelerate the programmer has no control over how to 
decompose his computation onto the GPU or how to make use of shared memory 
resources. For many users, remaining entirely within Haskell will be
a big attraction of Accelerate.

@@

The Copperhead \cite{Copperhead} system compiles a subset of Python to run 
on GPUs. Much like other languages mentioned here, Copperhead identifies 
usages of certain parallel primitives that can be executed in parallel 
on the GPU (such as reduce, scan and map). But Copperhead also 
allows the expression of nested data-parallelism and is in that 
way different from both Accelerate and Obsidian.

@@

In the implementation of Obsidian we use a very simple and direct approach 
to monad reification. Related to this work is the approach based on a 
continuation monad by Persson et al~\cite{Generic}. Recent work by Sculthorpe et al. in reference ~\cite{NEIL} is also related. 
Our approach to monad reification is simpler but less general.

% ---------------------------------------------------------------------------
% FUTURE WORK 
% --------------------------------------------------------------------------- 
\section{Future Work} 
\label{sec:Future}
%\emph{*** TEXT LEADING INTO THE THREE SUBSECTIONS}

\subsection{Increase low level control}

@@

\subsection{To automate or not to automate} 
Obsidian makes use of types that model the GPU hierarchy. This limits 
the kind of programs one can express but also ensures that what we can write 
is suitable for the GPU. An alternative is to remove these types and at the 
same time make the compiler more clever. This would move some of the 
required GPU detail knowledge from the programmer into compiler transformations. 
We have recently started a project to investigate this change of Obsidian
together with a Master's student.

\subsection{To optimise or not to optimise} 
The version of Obsidian described here does not try to use any compiler optimisations techniques. Instead, we are expecting that the 
CUDA compiler will apply a good set of techniques, from common subexpression 
elimination to more GPU specific transformations. As future work, we 
will investigate to what extent this reliance on the CUDA
compiler is justified. Also, we want to 
see if there are high-level optimisations that we can perform more easily 
on our intermediate representations of GPU kernels than what is possible 
on low-level CUDA code.

\subsection{Kernel Coordination}

@@

% ---------------------------------------------------------------------------
% CONCLUSION
% --------------------------------------------------------------------------- 
\section{Conclusion}

@@

Obsidian lends itself well to the kind of experimentation with low level 
GPU details that allow for the implementation of efficient kernels. This is 
illustrated in section \ref{sec:redOpt}, where we manage to apply techniques
used in the NVIDIA tutorial ~\cite{RedOpt}. The case study also show hows we 
can compose kernels and thus reuse prior effort. 
%For example, {\tt red4} 
%to {\tt red7} all make use of {\tt red3} for their parallel phases.

@@

% ---------------------------------------------------------------------------
% ACKNOWLEDGEMENTS
% --------------------------------------------------------------------------- 
\acks
The work on monad reification presented in this paper is joint 
work between Josef Svenningsson and Joel Svensson.

@@

\bibliographystyle{abbrvnat}
\bibliography{paper}
% The bibliography should be embedded for final submission.

@@

\end{document}

>>>>

%\documentclass[preprint]{sigplanconf}

@@

%\usepackage{amsmath}
%\usepackage{graphicx}
%\usepackage{placeins}
%\usepackage{tikz}
%\usepackage{pgfplots}
%\usetikzlibrary{shapes,snakes,positioning}
%\usepackage{fancyvrb}
%\usepackage[T1]{fontenc}
%\newcounter{arraycard}

%\begin{document}

%\conferenceinfo{WXYZ '05}{date, City.} 
%\copyrightyear{2005} 
%\copyrightdata{[to be supplied]}

%\titlebanner{DRAFT}        % These are ignored unless
%\preprintfooter{Obsidian: GPU kernel generation}   % 'preprint' option specified.

%\title{Embedded Language for High Performance GPU Computing}
%\title{Towards An Embedded Language\\ for High Performance GPU Computing}  
%\title{An Embedded Language for Resource Aware \\ GPU Kernel Programming}  
%\title{A High-Level Embedded Language for Low-Level GPU Kernel Programming}

@@

%\authorinfo{Bo Joel Svensson \and Mary Sheeran} 
%           {Chalmers University of Technology}
%           {\{joels,ms\} at chalmers dot se}

\newcommand{\subsubsubsection}[1]{\paragraph{#1:}}
%% \vspace{1ex}

%% \noindent\emph{\bf#1}\newline

%% \vspace{1ex}
%% \noindent
%% }

%\maketitle
% ---------------------------------------------------------------------------
%
% --------------------------------------------------------------------------- 
\subsection*{abstract}
%\begin{abstract}

@@

%\end{abstract}

%\category{CR-number}{subcategory}{third-level}
%\category{D.3.2}{Programming Languages}{Language Classifications}[Applicative (functional) languages; Concurrent, distributed, and parallel languages]
%\category{D.3.4}{Programming Languages}{Processors}[Code generation]

%\terms
%term1, term2
%\terms
%Languages, Performance

%\keywords
%keyword1, keyword2
%\keywords
%Data parallelism, array programming, embedded language

% ---------------------------------------------------------------------------
% INTRODUCTION 
% --------------------------------------------------------------------------- 
\subsection{Introduction}

@@

% ---------------------------------------------------------------------------
% Motivation and Contributions
% --------------------------------------------------------------------------- 
\subsubsection{Motivation and contributions}

@@

% ---------------------------------------------------------------------------
% The GPU 
% --------------------------------------------------------------------------- 
\subsubsection{The GPU and CUDA}
\label{sec:GPU}

@@

% ---------------------------------------------------------------------------
% PROGRAMMING IN OBSIDIAN 
% --------------------------------------------------------------------------- 
\subsection{A taste of Obsidian programming}

@@

There are two different array representations, {\em Pull} and 
{\em Push} arrays. A pull array naturally represents gather operations,
while a push array captures scattering operations. Reference \cite{Obsidian-Expressive} 
describes the introduction of push arrays into Obsidian.

@@

% ---------------------------------------------------------------------------
% OBSIDIAN Internals 
% --------------------------------------------------------------------------- 
\subsection{Obsidian internals}

@@

% ---------------------------------------------------------------------------
% Elements
% --------------------------------------------------------------------------- 
\subsubsection{Scalars}

@@

% ---------------------------------------------------------------------------
% PUSH AND PULL ARRAYS 
% --------------------------------------------------------------------------- 
\subsubsection{Pull arrays}

@@

Very similar representations are used 
in many other embedded languages. The 
embedded language Pan ~\cite{COMPILEEDSL} used a similar representation for 
images and it has been used since in Obsidian, Feldspar ~\cite{FELDSPAR} 
and Repa ~\cite{REPA}. In Obsidian, there is a {\tt force} function that,
when applied to a pull array, writes the elements of that array 
into shared memory using as many threads as the pull array is long. Hence, 
the {\tt force} function can be used to express parallelism in Obsidian.

@@

% ---------------------------------------------------------------------------
% Thread, Block and Grid Computations 
% --------------------------------------------------------------------------- 
\subsubsection{Thread, Block and Grid Computations} 
\label{sec:Program}

The programs that can be specified in Obsidian
are represented in a {\tt Program} data type.
This data type also represents the lowest level programming interface 
exposed by Obsidian. In reference ~\cite{JOSEFJOELCSORT} we make use of these low-level 
capabilities to implement histograms using atomic operations (not shown here).

@@

% ---------------------------------------------------------------------------
% Push Arrays
% --------------------------------------------------------------------------- 
\subsubsection{Push arrays} 
\label{sec:PushArrays}

@@

% ---------------------------------------------------------------------------
% Interplay 
% --------------------------------------------------------------------------- 
\subsubsection{Push and pull array interplay} 
\label{sec:interplay}

@@

% ---------------------------------------------------------------------------
% Compilation to CUDA 
% --------------------------------------------------------------------------- 
\subsubsection{Compilation to CUDA}
\label{sec:Compile}

@@

% ---------------------------------------------------------------------------
% More example programs 
% --------------------------------------------------------------------------- 
\subsection{Case studies} 
\label{sec:CASESTUDIES}

@@

%% ------------------------------------------------------------------------
%% Reduction
%% ------------------------------------------------------------------------
\subsubsection{Reduction}  
\label{sec:redOpt}

\input{./hl/reductionfigs.tex}

@@

Some of the reduction 
kernels will also require that the operation is commutative. 
%A typical reduction is to sum up the elements of an array, here this is used 
%as the running example. In this section we implement a series of reduction 
%kernels. 
To illustrate the kind of low level control that an 
Obsidian programmer has over expressing details of a kernel, we show a series 
of reduction kernels, each with different optimisations applied. Many of the 
optimisations that are applied to the kernels are influenced by a 
tutorial presentation from NVIDIA~\cite{reduction}.

@@

\subsubsubsection{Reduction 1}

@@

\subsubsubsection{Reduction 2}

@@

\subsubsubsection{Reduction 3}

@@

\subsubsubsection{Reduction 4}

@@

\subsubsubsection{Reduction 5}

@@

\subsubsubsection{Reductions 6 and 7}

@@

\subsubsubsection{Larger local reductions} 
Figure \ref{fig:reducetablel} shows running time results for kernels that are 
very similar to the ones described above, but with a larger number 
of elements per local reduction (that is per block). This means that all the kernels need 
to introduce a small amount of sequentiality at the beginning. This is 
done by introducing a new force function:

@@

\subsubsection{Scan}

@@

\input{./hl/scanfigs.tex}

@@

\subsubsubsection{Two elements per thread}

@@

Both to avoid conditionals and to allow
for larger scans per block, we move to two elements per thread.
Each phase of the algorithm is a parallel for loop that is executed 
by half as many threads as there are elements to scan. The body of the loop
performs one operation and one copy, using bit-twiddling to compute indices.
Note the use of two {\em write functions} in sequence.
Similar patterns were used in our implementations of sorting networks~\cite{Obsidian-Expressive}, for similar reasons.

@@

\subsection{Combining kernels to solve large problems} 
\label{sec:Benchmarks}

@@

\subsubsection{Large reductions}

@@

% ---------------------------------------------------------------------------
% RELATED WORK 
% --------------------------------------------------------------------------- 
\subsubsection{Related work}

@@

While remaining in the imperative world, but going all the way to the 
other end of the high-level - low-level spectrum, we have the NVIDIA
Thrust Library~\cite{THRUST}.
Thrust offers a programming model where details of GPU architecture are 
completely abstracted away. Here, the programmer expresses algorithms 
using building blocks like: \emph{Sort}, \emph{Scan} and \emph{Reduce}.

Data.Array.Accelerate is a language embedded in Haskell for GPU programming 
~\cite{ACCELERATEDAMP11}. The abstraction level is comparable to that of Thrust. In 
other words, Accelerate hides most GPU details from the programmer. Accelerate 
provides a set of operations (that are parallel and suitable for GPU execution, 
much like in Thrust) implemented as skeletons. Recent work has permitted
the optimisation of Accelerate programs using fusion techniques to decrease
the number of kernel invocations needed (see reference ~\cite{ACCOPT}).
%Accelerate programs are being optimised using using fusion 
%techniques to decrease the number of kernel invocations needed. 
It seems to us that when using Accelerate the programmer has no control over how to 
decompose his computation onto the GPU or how to make use of shared memory 
resources. For many users, remaining entirely within Haskell will be
a big attraction of Accelerate.

@@

The Copperhead \cite{copperhead} system compiles a subset of Python to run 
on GPUs. Much like other languages mentioned here, Copperhead identifies 
usages of certain parallel primitives that can be executed in parallel 
on the GPU (such as reduce, scan and map). But Copperhead also 
allows the expression of nested data-parallelism and is in that 
way different from both Accelerate and Obsidian.

@@

In the implementation of Obsidian we use a very simple and direct approach 
to monad reification. Related to this work is the approach based on a 
continuation monad by Persson et al~\cite{Generic}. Recent work by Sculthorpe et al. in reference ~\cite{sculthorpe2013constrained} is also related. 
Our approach to monad reification is simpler but less general.

% ---------------------------------------------------------------------------
% FUTURE WORK 
% --------------------------------------------------------------------------- 
\subsection{Future Work} 
\label{sec:Future}
%\emph{*** TEXT LEADING INTO THE THREE SUBSECTIONS}

\subsubsection{Increase low level control}

@@

\subsubsection{To automate or not to automate} 
Obsidian makes use of types that model the GPU hierarchy. This limits 
the kind of programs one can express but also ensures that what we can write 
is suitable for the GPU. An alternative is to remove these types and at the 
same time make the compiler more clever. This would move some of the 
required GPU detail knowledge from the programmer into compiler transformations. 
We have recently started a project to investigate this change of Obsidian
together with a Master's student.

\subsubsection{To optimise or not to optimise} 
The version of Obsidian described here does not try to use any compiler optimisations techniques. Instead, we are expecting that the 
CUDA compiler will apply a good set of techniques, from common subexpression 
elimination to more GPU specific transformations. As future work, we 
will investigate to what extent this reliance on the CUDA
compiler is justified. Also, we want to 
see if there are high-level optimisations that we can perform more easily 
on our intermediate representations of GPU kernels than what is possible 
on low-level CUDA code.

\subsubsection{Kernel Coordination}

@@

% ---------------------------------------------------------------------------
% CONCLUSION
% --------------------------------------------------------------------------- 
\subsection{Conclusion}

@@

Obsidian lends itself well to the kind of experimentation with low level 
GPU details that allow for the implementation of efficient kernels. This is 
illustrated in section \ref{sec:redOpt}, where we manage to apply techniques
used in the NVIDIA tutorial ~\cite{reduction}. The case study also show hows we 
can compose kernels and thus reuse prior effort. 
%For example, {\tt red4} 
%to {\tt red7} all make use of {\tt red3} for their parallel phases.

@@

% ---------------------------------------------------------------------------
% ACKNOWLEDGEMENTS
% --------------------------------------------------------------------------- 
\subsection*{Acknowledgments}
The work on monad reification presented in this paper is joint 
work between Josef Svenningsson and Joel Svensson.

@@

%\bibliographystyle{abbrvnat}
%\bibliography{paper}
% The bibliography should be embedded for final submission.

@@

%\end{document}

### 2013-10-02 src2/ifl/GPUs.tex
apa

\section{Graphics Processing Units, accessible high performance parallel computing}
\label{sec:gpu}

@@

See reference~\cite{GPUComputing} for information about
the very similar AMD GPU architecture. We have used the NVIDIA architecture,
but developments are similar at AMD. Intel's Larrabee processor points to a
future in which each individual core is considerably more powerful than in
today's GPUs~\cite{IntelLarrabee}.

>>>>

\subsection{Graphics Processing Units, accessible high performance parallel computing}
\label{sec:gpu}

@@

See reference~\cite{GPUComputing} for information about
the very similar AMD GPU architecture. We have used the NVIDIA architecture,
but developments are similar at AMD. Intel's Larrabee processor points to a
future in which each individual core is considerably more powerful than in
today's GPUs~\cite{larrabee}.

### 2013-10-02 src2/ifl/combinators.tex
apa

\section{Connection patterns for hardware design and parallel programming}\label{sec:combinators}

@@

For example, an important pattern is parallel prefix or scan. Given inputs $[x_0, x_1 \ldots x_{n-1}]$, the prefix problem is to compute each $x_0 \circ x_1 \circ \ldots \circ x_j$ for $0 \leq j < n$, for $\circ$ an associative, but not necessarily commutative, operator. For example, the prefix sum of {\tt [1..10]} is {\tt [1,3,6,10,15,21,28,36,45,55]}. There is an obvious sequential solution, but in circuit design one is often aiming for a circuit that exploits parallelism, and so is faster (but also larger). In a construction attributed to Sklansky, one can perform the prefix
calculation by first, recursively, performing the prefix calculation on 
each half of the input, and then combining (via the operator) the last output of the first of these
recursive calls with each of the outputs of the second.
For instance, to calculate the prefix sum of {\tt [1..10]}, one can compute the prefix
sums of {\tt [1..5]} and {\tt [6..10]}, giving {\tt [1,3,6,10,15]}
and {\tt [6,13,21,30,40]}, respectively. 
The final step is to add the last element of the output of the first recursive call ({\tt 15}) to each element of
the output of the second.

To express the construction in Lava, we make use of two connection patterns.
{\tt two :: ([a] -> [b]) -> [a] -> [b]} applies its component to
the top and bottom halves of the input list, concatenating the two sub-lists
that result from these applications.
Thus, {\tt two (sklansky plus)} applied to {\tt [1..10]} gives {\tt [1,3,6,10,15,6,13,21,30,40]}.
Left-to-right serial composition has type
{\tt (a -> b) -> (b -> c) -> a -> c} and is written as infix {\tt ->-}.
The description of the construction mixes the use of connection patterns, giving a form of reuse, with the naming of ``wires''.

>>>>

\subsection{Connection patterns for hardware design and parallel programming}\label{sec:combinators}

@@

For example, an important pattern is parallel prefix or scan. Given inputs\newline $[x_0, x_1 \ldots x_{n-1}]$, the prefix problem is to compute each $x_0 \circ x_1 \circ \ldots \circ x_j$ for $0 \leq j < n$, for $\circ$ an associative, but not necessarily commutative, operator. For example, the prefix sum of {\tt [1..10]} is {\tt [1,3,6,10,15,21,28,36,45,55]}. There is an obvious sequential solution, but in circuit design one is often aiming for a circuit that exploits parallelism, and so is faster (but also larger). In a construction attributed to Sklansky, one can perform the prefix
calculation by first, recursively, performing the prefix calculation on 
each half of the input, and then combining (via the operator) the last output of the first of these
recursive calls with each of the outputs of the second.
For instance, to calculate the prefix sum of {\tt [1..10]}, one can compute the prefix
sums of {\tt [1..5]} and {\tt [6..10]}, giving {\tt [1,3,6,10,15]}
and {\tt [6,13,21,30,40]}, respectively. 
The final step is to add the last element of the output of the first recursive call ({\tt 15}) to each element of
the output of the second.

To express the construction in Lava, we make use of two connection patterns.\newline
{\tt two :: ([a] -> [b]) -> [a] -> [b]} applies its component to
the top and bottom halves of the input list, concatenating the two sub-lists
that result from these applications.
Thus, {\tt two (sklansky plus)} applied to {\tt [1..10]} gives {\tt [1,3,6,} {\tt10,15,6,13,21,30,40]}.
Left-to-right serial composition has type
{\tt (a -> b) -> (b -> c) -> a -> c} and is written as infix {\tt ->-}.
The description of the construction mixes the use of connection patterns, giving a form of reuse, with the naming of ``wires''.

### 2013-10-02 src2/ifl/obsidian.tex
apa

% ------------------------------------------------------------------------------
\section{Obsidian: a domain specific embedded language for GPU programming}
\label{sec:obsidian}

@@

% ------------------------------------------------------------------------------
\subsection{Implementation}

@@

The above expressions are dynamic in the sense that they can be used to 
represent values of {\tt Int}, {\tt Float} and {\tt Bool} type. 
This follows the approach from Compiling Embedded Languages 
\cite{Elliott03:CompileDSEL-JFP}. However, 
to obtain a typed environment in which to operate, phantom types are used.

@@

At the code generation phase this function is evaluated using a variable 
representing a thread Id. The result is an expression looking as follows:
\begin {code}
E (BinOp Add (Index (Variable "input") (Variable "tid")) (LitInt 1))
\end{code}

>>>>

% ------------------------------------------------------------------------------
\subsection{Obsidian: a domain specific embedded language for GPU programming}
\label{sec:obsidian}

@@

% ------------------------------------------------------------------------------
\subsubsection{Implementation}

@@

The above expressions are dynamic in the sense that they can be used to 
represent values of {\tt Int}, {\tt Float} and {\tt Bool} type. 
This follows the approach from Compiling Embedded Languages 
\cite{COMPILEEDSL}. However, 
to obtain a typed environment in which to operate, phantom types are used.

@@

At the code generation phase this function is evaluated using a variable 
representing a thread Id. The result is an expression looking as follows:
\begin{code}
E (BinOp Add (Index (Variable "input") (Variable "tid")) 
             (LitInt 1))
\end{code}
